\vspace{-1.5cm} Unsupervised learning of probabilistic programs with latent predicate networks

### 2014-11-07 drafts/nips_workshop/nips_workshop.tex
Rewrote introduction; fixed figure name.

\title{Latent Predicate Networks: unsupervised learning of  probabilistic programs with context sensitive grammars}

@@

\begin{document}

\maketitle

\begin{abstract}

@@

Probabilistic programs can successfully model many aspects of human
cognition [CITIATONS]. Part of their strength as compared to other
models of cognition lies in their ability to flexibly integrate both
symbolic and statistical information within a single framework.

One area probabilistic programming systems lag behind human cognition
is in their ability to automatically induce entire

Although concept learning and language acquisition have typically been
treated as distinct problems in AI, linguistics and cognitive
development, they are strongly coupled for a child learning to
understand language. More generally, people learn many abstract
concepts primarily through language even though understanding language
depends on understanding the underlying concepts. Research in concept
learning is often focused on concepts grounded in perceptual features,
and while it is almost certainly true that many concepts are learned
via generalization from concrete examples, some concepts cannot be
learned this way.

Number concepts are a good example: children do not learn about the
meaning of ``seventy five'' by seeing examples of seventy five things;
they do not know that ``seventy five'' is more than ``twenty five''
because of their perceptual experiences of these quantities. Rather,
children learn the meaning of ``seventy five'' (or ``a billion and
five'') by noticing how number words are used in language, in counting
sequences, in arithmetic exercises, {\it etc}. Other good examples of such
abstract concepts are kinship and social relations ({\it e.g.} ``my father
in law's grandmother''), temporal relations (``the day after last
Thanksgiving.''), and spatial relations (``above and just to the left
of'').

Such concepts share many of the properties of language syntax: they
are unbounded in number, they derive their meanings via composition,
and, although people only ever say, hear, or read about a small number
of them, they are able to reason correctly about any them. There seems
to be a grammar to these concepts, and grasping this grammar is
critical to understanding their meanings and how to use them. This
motivates our approach here, which is to apply the tools of
probabilistic grammars more familiar from studies of syntax to the
problem of concept aquisition.

{\bf include somewhere:} A general probabilistic logic
programming system based on PROLOG, PRISM provides built-in predicates
for probabilistic execution and Bayesian inference over logic programs
with stochastic choices.

Doing so requires overcoming some technical barriers.

First, many formal systems of probabilistic programming are so
expressive as to make learning generally intractable. To address this,
we use Range Concatenation Grammar, a context-sensitive grammar
formalism originally developed in linguistics and extend this to a
probabilistic model of programs.

Second, the categories of syntax -- the nonterminals of the grammar --
are often assumed to be known to children innately and given to
automated learners by human experts. The categories that underlie
conceptual knowledge, on the other hand, are far more numerous, vary
from domain to domain, and are unlikely to be known to the
learner. This motivates our use of latent predicates that, through
learning, assume the role of a domain's underlying concepts (in the
number domain, these might correspond to the concepts of
successorship, order of magnitude, magnitude comparison, exact
vs. approximate, {\it etc}.).

Finally, inducing probabilistic context-sensitive grammars with latent
predicates threatens to be intractable: our goal is to find a middle
ground between expressivity and tractability. Using
PRISM~\cite{DBLP:journals/jair/SatoK01} -- a probabilistic logic
programming system that naturally implements efficient dynamic
programming algorithms for our models -- we are able to explore which
domains and which grammar architectures are a good fit for this
grammar-based approach.

The rest of this paper describes early work on this approach. First,
we present Latent Predicate Networks (LPNs), our probabilistic model
of concept learning. Then, we describe our approach to inference and
its implementation in PRISM. Finally, we present preliminary
experimental results in the domain of number concept learning that
demonstrate the promise of this approach.

>>>>

\title{\vspace{-1.5cm} Unsupervised learning of probabilistic programs with latent predicate networks}

@@

\begin{document}
\maketitle

\vspace{-1cm}
\begin{abstract}

@@

Probabilistic programs combine symbolic and statistical knowledge in way that is
natural for humans to express. In fact, probabilistic programs have
been proposed as models of human cognition in a variety of domains
[CITATION]. As a suitable representation for learning, however, the flexibility
of probabilistic programs poses a challenge. At one extreme, we can
approach learning as fixing the symbolic structure of a probabilistic
program and fitting some of its parameters; at the other extreme, the
structure is unspecified and learning is performed by searching over
valid syntactic forms. The former sacrifices the symbolic flexibility
that probabilistic programs have to offer, and the latter is
computationally infeasible. It is natural to suppose, then, that human
learning charts a middle course between these extremes, restricting
the space of probabilistic programs -- perhaps in a domain-specific
way -- to accomodate both flexibility and tractability.

Here, we present statistical learning (using the PRISM probabilistic
logic programmming language) of a restricted type of probabilistic
program, dubbed a \emph{latent predicate network} (LPN), for learning
problems at the intersection of concept learning and language
acquisition. People learn many abstract concepts primarily through
language, and understanding language depends on understanding the
underlying concepts. Number concepts are a good example: children do
not learn about the meaning of ``seventy five'' by seeing examples of
seventy five things; they do not know that ``seventy five'' is more
than ``twenty five'' because of their perceptual experiences of these
quantities. Rather, children learn the meaning of ``seventy five'' (or
``a billion and five'') by noticing how number words are used in
language, in counting sequences, in arithmetic exercises, {\it
  etc}. Other good examples of such abstract concepts are kinship and
social relations ({\it e.g.} ``my father in law's grandmother''),
temporal relations (``the day after last Thanksgiving.''), and spatial
relations (``above and just to the left of'').

We use LPNs in an attempt to address the issues of tractibility and
symbolic flexibility. We address the issue of tractability by
extending Range Concatenation Grammars -- a context-sensitive grammar
formalism originally developed in linguistic -- with a probabilistic
interpretation, resulting in probabilistic programs in which
inference, and thus learning, is relatively tractable. We address the
issue of structural flexibility by building in latent predicates,
which, through learning, assume the role of a domains underlying concepts (in the
number domain, these might correspond to the concepts of
successorship, order of magnitude, magnitude comparison, exact
vs. approximate, {\it etc}.).

% Such concepts share many of the properties of language syntax: they
% are unbounded in number, they derive their meanings via composition,
% and, although people only ever say, hear, or read about a small number
% of them, they are able to reason correctly about any them. There seems
% to be a grammar to these concepts, and grasping this grammar is
% critical to understanding their meanings and how to use them. This
% motivates our approach here, which is to apply the tools of
% probabilistic grammars more familiar from studies of syntax to the
% problem of concept aquisition.

% {\bf include somewhere:} A general probabilistic logic
% programming system based on PROLOG, PRISM provides built-in predicates
% for probabilistic execution and Bayesian inference over logic programs
% with stochastic choices.

% Doing so requires overcoming some technical barriers.

% First, many formal systems of probabilistic programming are so
% expressive as to make learning generally intractable. To address this,
% we use Range Concatenation Grammar, a context-sensitive grammar
% formalism originally developed in linguistics and extend this to a
% probabilistic model of programs.

% Second, the categories of syntax -- the nonterminals of the grammar --
% are often assumed to be known to children innately and given to
% automated learners by human experts. The categories that underlie
% conceptual knowledge, on the other hand, are far more numerous, vary
% from domain to domain, and are unlikely to be known to the
% learner. This motivates our use of latent predicates that, through
% learning, assume the role of a domain's underlying concepts (in the
% number domain, these might correspond to the concepts of
% successorship, order of magnitude, magnitude comparison, exact
% vs. approximate, {\it etc}).

% Finally, inducing probabilistic context-sensitive grammars with latent
% predicates threatens to be intractable: our goal is to find a middle
% ground between expressivity and tractability. Using
% PRISM~\cite{DBLP:journals/jair/SatoK01} -- a probabilistic logic
% programming system that naturally implements efficient dynamic
% programming algorithms for our models -- we are able to explore which
% domains and which grammar architectures are a good fit for this
% grammar-based approach.

The rest of this paper describes early work on this approach. First,
we present Latent Predicate Networks (LPNs). Then, we describe our
approach to inference and its implementation in PRISM. Finally, we
present preliminary experimental results in the domain of number
concept learning.
